How does the writing system get informative? Differentiation or conservation?

### 2023-06-22 manuscript/wp4.tex
ms

The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two spoken forms merge and become homophonous, but their heterographic spellings are conserved in the orthography. For example, the words \textit{night} and \textit{knight}---heterographic homophones in modern English---were previously pronounced differently from each other (/niçt/ and /kniçt/; in Old and Middle English, /kn/ was a permissible consonant cluster, as it still is in modern Dutch and German).

@@

French spellings
double letters inn
add -e

@@

>>>>

The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two (or more) spoken forms merge and become homophonous, but the original heterographic spellings are conserved in the orthography. In English, for example, the meat--meet merger, which was complete by around 1700, resulted in Middle English /ɛ:/ (spelled <ea>) and /e:/ (spelled <ee>) merging into /i:/ in Early Modern English, although the spellings were never updated in response, thus giving rise to a set heterographic homophones that persist in present-day English: \textit{heal}--\textit{heel}, \textit{leak}--\textit{leek}, \textit{meat}--\textit{meet}, \textit{read}--\textit{reed}, \textit{sea}--\textit{see}, \textit{team}--\textit{teem}, \textit{weak}--\textit{week}, etc. Similarly, other historic sound changes have likely resulted in (or contributed to) pairs of words entering a state of heterography, such as the reduction of the consonant cluster /kn/ into /n/ (\textit{knight}--\textit{night}, \textit{know}--\textit{no}, \textit{knot}--\textit{not}, etc.), the loss of the palatal fricative /ç/ (\textit{eight}--\textit{ate}, \textit{right}--\textit{rite}, \textit{sight}--\textit{site}, etc.), and the merger of the voiceless labial-velar fricative /ʍ/ into the voiced approximant /w/ (\textit{whale}--\textit{wail}, \textit{which}--\textit{witch}, \textit{whine}--\textit{wine}, etc.).

@@

French spellings
double letters in-inn, be-bee, for-fore-four, to-too-two, but-butt, by-bye-buy, more-moor, none-nun, we-wee, or-oar-ore, so-sew
add -e

@@

Why do we get redundant systems?

- lots of variation available, why condition it on shape and not color?
- strong cue from the spoken language hints that the language does not mark color, so the natural thing to think is that the spelling variation is based on shape

Why do we not get expressive systems?

- alignment problem - hard to align with someone when you can only send one word, not meta conversation
- strong signal from the spoken language that it doesn't mark color, peolpe believe spelling shoud represent the truth of speech
- social pressure to stick to the taught system - affraid to get in trouble or get partner in trouble
- the goal is perceived as reproduction rather than invention
